Patterns of floral and chromosome evolution in Onagraceae: a fossil-calibrated supermatrix analysis

### 2014-05-16 manuscript.tex
revisions and discussion

\begin{abstract}
Onagraceae is a cosmopolitan plant family in the order Myrtales comprised of 22 genera and 657 species.
I used a GenBank data-mining approach to construct an 11 gene supermatrix of 521 taxa in Onagraceae and Lythraceae.
Maximum likelihood and Bayesian inference phylogenetic analyses were performed, 
and divergence time estimates were calibrated using 5 fossils.
Chromosome number evolution was inferred using probabilistic models,
and ancestral states of petal color and petal number were reconstructed using maximum likelihood.
Correlated evolution between petal color and petal number was tested using
Bayesian stochastic character mapping.
The monophyly of all major clades in Onagraceae was supported,
and Onagraceae was estimated to have diverged from Lythraceae 109 Mya. 
The base chromosome number of Onagraceae was inferred to be $x=5$.
Petal color and petal number were found
to follow a pattern of correlated evolution ($p=0.00$), 
suggesting concerted shifts in floral traits may play an important role in shaping diversification of Onagraceae. 
Further analyses should be performed to test whether shifts in
floral traits are correlated with pollinator shifts,
and to determine how these shifts affect diversification rates.
Furthermore, this work should be extended to test historical biogeographical hypotheses
that explain the intercontinental disjunct distribution of Onagraceae.
\end{abstract}

@@

In pursuit of large-scale evolutionary questions,
biologists are increasingly using massive phylogenetic
datasets to reconstruct ever-growing portions of the tree of life. 
A common approach to building large phylogenetic trees has been
the supermatrix approach, in which multiple datasets are
combined and analyzed simultaneously \citep{DeQueiroz2007}.
However, assembling and utilizing supermatrices is not trivial
due to challenges such as determining homology of molecular
sequences, assembling chimeric operational taxonomic units, 
limiting the amount of missing data,
and managing data integrity and computational feasibility \citep{Roure2007, Yan2005}.
Despite these difficulties, considerable bioinformatic advances 
have made supermatrix construction more common,
particularly in large-scale evolutionary plant studies.
For example, \citet{Smith2011}
inferred a phylogeny of 55473 seed plants,
\citet{Hinchliff2013} estimated the phylogeny of 1500 members of Cyperaceae,
and \citet{Sarkinen2013} built a time calibrated tree of Solanaceae with 1076 tips.

With over 5000 DNA sequences available in GenBank from previous studies, 
the plant family Onagraceae is an
ideal group to apply a data-mining supermatrix approach.
Onagraceae, the evening primrose family, is a well-studied 
monophyletic clade
of flowering plants with 22 described genera and around 660 species \citep{wagner2007revised}.
The family is found on all continents except Antartica,
and pollen fossils date Onagraceae to the Upper Cretaceous \citep{grimsson}.
Onagraceae falls within the order Myrtales,
and previous molecular phylogenetic studies find it most closely related to Lythraceae \citep{conti1997, Sytsma2004}.
Though there have been many genus-level molecular phylogenetic studies in Onagraceae
\citep{Berry2004, Evans2009, Hoggard2004, Xie2009, Baum1994, Wagner2005}, 
the most recent family-level analyses \citep{Levin2003, Levin2004}
used only one or two exemplars from some genera. 
Furthermore, the historical biogeography, divergence times of the major clades, and 
family-level patterns and processes of diversification remain poorly understood
and untested over a molecular phylogeny.

Here, I use a well-resolved and densely-sampled supermatrix phylogeny of Onagraceae to
test whether the taxonomic groups described in \citet{wagner2007revised} are monophyletic,
and to answer the following questions:
1) When did the major clades diverge?
2) What is the pattern of chromosome number evolution?
3) What are the patterns of petal color and petal number evolution, and are their 
evolution correlated?
Lastly, I consider ways in which these analyses can be extended
to further examine the evolution of Onagraceae, 
and discuss the novel method used here to construct the supermatrix.

@@

\paragraph{Phylogeny and divergence time estimates}
The topologies of the ML and Bayesian phylogenies were nearly identical for all major clades within Onagraceae, 
so only the Bayesian MCC tree (Figures \ref{posteriors} and \ref{genera}) is shown here.
The major exception is that in the ML tree the genus \textit{Chamerion} was recovered
as sister to \textit{Epilobium}, whereas in the MCC tree XXXXXX.......
In the MCC tree all Onagraceae genera described in \citet{wagner2007revised} were recovered as monophyletic clades with posterior probabilities $> 0.95$
except for sister genera \textit{Neoholmgrenia} and \textit{Camissoniopsis} (posterior = 0.31) (Figure \ref{posteriors}).
Onagraceae was found to diverge from Lythraceae at 109 Mya (Figure \ref{genera}). 
Divergence time estimates of other major clades and 95\% highest posterior density (HPD) intervals can be seen in Table \ref{times}.

@@

\paragraph{Floral evolution}
The maximum likelihood estimate of petal color for the 
last common ancestor of Onagraceae was pink 
with a proportional likelihood ($PL$) of $0.56$ (Figure \ref{petal_color}).
White was the second most likely estimate ($PL=0.19$).
The maximum likelihood estimate of petal number for the
last common ancestor of Onagraceae was 4 ($PL=1.0$), with 8
independent transitions to 0, 2, or 5 petals (Figure \ref{petal_number}).
Petal color and petal number were found to significantly covary over the 
phylogeny ($D=0.263, p=0.00$). Multiple signficant correlations
($d_{ij}$) were found between individual states of each character
and are summarized in Table \ref{correlations}.

@@

discuss monophyly of tribe Epilobieae, which was supported in 
maximum likelihood analysis but not supported in Bayesian MCC, could be due to
fossil constraints on genus Epilobium, it may be that the fossil described should
have constrained Epilobium + Chamerion due to taxonomic changes.....
need to check fossil synapomorphies to see where it should be placed!

Petal color and petal number were found to follow a pattern 
of correlated evolution (p = 0.00), 
suggesting concerted shifts in floral traits 
may play an important role in shaping diversification of Onagraceae. 
Further analyses should be performed to test 
whether shifts in floral traits are correlated with pollinator shifts, 
and to determine how these shifts affect diversification rates. 
Furthermore, this work should be extended to test historical 
biogeographical hypotheses that explain the intercontinental disjunct distribution of Onagraceae.

For example, the PhyLoTA Browser 
provides a web interface to view all GenBank sequences within
taxonomic groups clustered into homologs \citep{Sanderson2008}.
A different approach is implemented in the program PHLAWD,
which constructs clusters of homologs around
guide sequences provided by the user \citep{Smith2009}.
However, the method used by SUMAC combines both approaches;
the user can perform an exploratory clustering of all GenBank sequences
within a taxonomic group, and also provide guide sequences to build
homologous clusters in a more targeted manner. SUMAC provides
a framework with which GenBank can be repeatedly mined 
using different taxonomic groups and/or different guide sequences
and the resulting data matrices can be compared.

describe some recently developed metrics that can
be used to assess the informativeness of phylogenetic supermatrices.

>>>>

\begin{abstract}
Onagraceae is a cosmopolitan plant family in the order Myrtales comprised of 22 genera and 657 species.
I used a GenBank data-mining approach to construct an 11 gene supermatrix of 521 taxa in Onagraceae and Lythraceae.
Maximum likelihood and Bayesian inference phylogenetic analyses were performed, 
and divergence time estimates were calibrated using 5 fossils.
Chromosome number evolution was inferred using probabilistic models,
and ancestral states of petal color and petal number were reconstructed using maximum likelihood.
Correlated evolution between petal color and petal number was tested using
Bayesian stochastic character mapping.
The monophyly of all genera in Onagraceae was supported,
and Onagraceae was estimated to have diverged from Lythraceae 109 Mya. 
The base chromosome number of Onagraceae was inferred to be $x=5$.
Petal color and petal number were found
to follow a pattern of correlated evolution ($p=0.00$), 
suggesting concerted shifts in floral traits may play an important role in shaping diversification of Onagraceae. 
Further analyses should be performed to test whether shifts in
floral traits are correlated with pollinator shifts,
and to determine how these shifts affect diversification rates.
Furthermore, this work should be extended to test historical biogeographical hypotheses
that explain the intercontinental disjunct distribution of Onagraceae.
\end{abstract}

@@

In pursuit of large-scale evolutionary questions,
biologists are increasingly using massive phylogenetic
datasets to reconstruct ever-growing portions of the tree of life. 
A common approach to building large phylogenetic trees has been
the supermatrix approach, in which multiple datasets are
combined and analyzed simultaneously \citep{DeQueiroz2007}.
However, assembling and utilizing supermatrices is challenging
due to difficulties such as determining homology of molecular
sequences, assembling chimeric operational taxonomic units, 
limiting the amount of missing data,
and managing data integrity and computational feasibility \citep{Roure2007, Yan2005}.
Despite these challenges, considerable bioinformatic advances 
have made supermatrix construction more common,
particularly in large-scale evolutionary plant studies.
For example, \citet{Smith2011}
inferred a phylogeny of 55473 seed plants,
\citet{Hinchliff2013} estimated the phylogeny of 1500 members of Cyperaceae,
and \citet{Sarkinen2013} built a time calibrated tree of Solanaceae with 1076 tips.

The plant family Onagraceae is an
ideal group to apply a data-mining supermatrix approach
due to the over 5000 DNA sequences available in GenBank from previous studies.
Onagraceae, the evening primrose family, is a well-studied 
monophyletic clade
of flowering plants with 22 described genera and around 660 species \citep{wagner2007revised}.
The family is found on all continents except Antartica,
and pollen fossils date Onagraceae to the Upper Cretaceous \citep{grimsson}.
Onagraceae falls within the order Myrtales,
and previous molecular phylogenetic studies find it most closely related to Lythraceae \citep{conti1997, Sytsma2004}.
Though there have been many genus-level molecular phylogenetic studies in Onagraceae
\citep{Berry2004, Evans2009, Hoggard2004, Xie2009, Baum1994, Wagner2005}, 
the most recent family-level analyses \citep{Levin2003, Levin2004}
used only one or two exemplars from some genera. 
Furthermore, the historical biogeography, divergence times of the major clades, and 
family-level patterns and processes of diversification remain poorly understood
and untested over a molecular phylogeny.

Here, I use a well-resolved and densely-sampled supermatrix phylogeny of Onagraceae to
test the monophyly of the taxonomic groups described in \citet{wagner2007revised}
and estimate when they diverged.
I then inferred the pattern of chromosome number evolution using probabilistic models and
examined the patterns of petal color and petal number evolution, 
including a correlation test of the evolution of these two floral traits.
Lastly, I consider ways in which these analyses can be extended
to further examine the evolution of Onagraceae, 
and discuss the novel method used here to construct the phylogenetic supermatrix.

@@

\paragraph{Phylogeny and divergence time estimates}
The topologies of the ML and Bayesian phylogenies were nearly identical for all major clades within Onagraceae, 
so only the Bayesian MCC tree (Figures \ref{posteriors} and \ref{genera}) is shown here.
The only exception is the tribe Epilobieae \textit{sensu} \citet{wagner2007revised}; 
the ML analysis recovered tribe Epilobieae as monophyletic, 
whereas according to the Bayesian estimate it is paraphyletic.
All Onagraceae genera described in \citet{wagner2007revised} were recovered as monophyletic clades 
in the Bayesian MCC with posterior probabilities $> 0.95$,
except for sister genera \textit{Neoholmgrenia} and \textit{Camissoniopsis} (posterior = 0.31) (Figure \ref{posteriors}).
Onagraceae was found to diverge from Lythraceae at 109 Mya (Figure \ref{genera}). 
Divergence time estimates of other major clades and 95\% highest posterior density (HPD) intervals can be seen in Table \ref{times}.

@@

\paragraph{Floral evolution}
The maximum likelihood estimate of petal color for the 
last common ancestor of Onagraceae was pink 
with a proportional likelihood ($PL$) of $0.56$ (Figure \ref{petal_color}).
White was the second most likely estimate ($PL=0.19$).
The maximum likelihood estimate of petal number for the
last common ancestor of Onagraceae was 4 ($PL=1.0$), with 8
independent transitions to 0, 2, or 5 petals (Figure \ref{petal_number}).
Petal color and petal number were found to significantly covary over the 
phylogeny ($D=0.263, p=0.00$). Multiple significant correlations
($d_{ij}$) were found between individual states of each character
and are summarized in Table \ref{correlations}.

@@

The paraphyly of tribe Epilobieae in the Bayesian MCC tree (Figure \ref{genera})
conflicts with both the 
maximum likelihood analysis (not shown) and the previous studies of \citet{Levin2003, Levin2004},
all of which found tribe Epilobieae to be monophyletic.
This incongruence is likely due to the placement of the
fossil constraint on the genus \textit{Epilobium} used in the Bayesian analysis. 
In this study I followed the taxonomy of \citet{wagner2007revised}, 
which splits tribe Epilobieae into the sister genera \textit{Epilobium}
and \textit{Chamerion}.
However, the pollen morphology descriptions used by \citet{grimsson} to identify the
\textit{Epilobium} fossil did not recognize \textit{Chamerion} as separate from
\textit{Epilobium} \citep{Punt2003, Patel1984}. 
The divergence times will be reestimated using the \textit{Epilobium} fossil 
to constrain both genera. 
This will likely result in topological changes to the Bayesian MCC, 
which means that the 
trait evolution estimates presented here should be considered preliminary.

The reconstruction of ancestral chromosome numbers 
validates many estimates made by \citet{Raven1979}.
Raven's estimates of the base chromosome number for the genera
\textit{Ludwigia}, \textit{Fuchsia}, \textit{Circaea}, \textit{Hauya}, 
\textit{Epilobium}, \textit{Camissonia}, and \textit{Oenothera}
all agree with the reconstruction presented here (Figure \ref{chromtree}).
However, for two genera I report estimates that differ from Raven's: 
for \textit{Clarkia} $x=9$ and for \textit{Lopezia} $x=10$.
Furthermore, \citet{Raven1979} proposed that the ancestral
chromosome number of all Onagraceae was $x=11$, whereas
my estimate suggests $x=5$. Further testing with the ChromEvol
models should be carried out, particularly to see if the inclusion
of the outgroup (Lythraceae) affected the estimate of the
base chromosome number of Onagraceae.

The analysis of petal color and petal number evolution using
Bayesian stochastic character mapping
revealed that the two floral traits followed a pattern 
of correlated evolution (p = 0.00), 
suggesting concerted shifts in floral traits 
may play an important role in shaping diversification of Onagraceae. 
To test this, further analyses will be performed to 
determine if these shifts in floral morphology affect diversification rates.
Other floral traits, such as corolla width and diurnal or nocturnal flower opening, 
could be included in future analyses.
Tests will also be performed to evaluate whether 
shifts in floral traits are correlated with pollinator shifts.
These analyses could provide evidence that pollinators are driving the
evolution of floral morphology in Onagraceae \citep{DeWittSmith2010}.

This work will also be extended to test historical 
biogeographical hypotheses that explain the intercontinental disjunct distribution of Onagraceae.
The family-level supermatrix phylogeny will be used to infer the location of the 
last common ancestor of Onagraceae, which has been hypothesized
to have been in South America \citep{raven1974angiosperm, raven1988onagraceae}.
The biogeographical analysis will also test the suggestion made by \citet{Berry2004} that \textit{Fuchsia}
originated in Central America. 
To my knowledge this will be the first reconstruction of 
the historical biogeography of Onagraceae over a molecular phylogeny.

The strength of the supermatrix approach to phylogenetic estimation
is that it combines data from diverse sources into one large analysis.
This strength is amplified by data-mining GenBank in an exploratory
fashion, so that sequence data not necessarily used in previous systematic
studies can also be incorporated into the supermatrix.
A number of software tools for building supermatrices are already available 
that use methods similar to those performed here.
For example, the PhyLoTA Browser 
provides a web interface to view all GenBank sequences within
taxonomic groups clustered into homologs \citep{Sanderson2008}.
A different approach is implemented in the program PHLAWD,
which constructs clusters of homologs around
guide sequences provided by the user \citep{Smith2009}.
However, the method used here (implemented in SUMAC) combines elements of both approaches;
the user can perform an exploratory clustering of all GenBank sequences
within a taxonomic group, and also provide guide sequences to build
homologous clusters in a more targeted manner. 
SUMAC is being extended to calculate metrics that assess
the informativeness of phylogenetic supermatrices 
that are unavailable in other software tools.
For example, the next version of SUMAC will calculate
an index of substitution saturation based on information entropy \citep{Xia2003}
for each alignment,
and also calculate the partial phylogenetic decisiveness 
of the supermatrix,
which measures the percentage of trees (out of all possible trees)
that can be inferred given the data missing from the supermatrix \citep{Sanderson2010, Fischer2012}.
By calculating these metrics SUMAC will provide a unique
framework with which GenBank can be repeatedly mined 
using different settings
and the resulting data matrices can be compared.
With further methodological refinement, 
supermatrix approaches will be indispensable
to future large-scale phylogenetic studies.
